Short-term neuronal and synaptic plasticity act in synergy for deviance detection in spiking networks

### 2023-01-30 paper-1/manuscript.tex
Editing

Having established that the increased response to A in \dev{} trials was driven by lower thresholds before stimulation, we next turned to the cause of this reduction. A priori, since threshold adaptation is a direct reflection of activity, and since the response to target trials was larger in \dev{} than in \msc{}, the lower thresholds in \dev{} must have been caused by lower activity in non-target trials, i.e., by lower responses to B in \dev{} than to B through E in \msc{}. Yet, since activity is modulated by thresholds as shown above, and since the dev sequence was random, implying that $\T[dev]{A}{} \approx \T[dev]{B}{}$, we might expect the lowered thresholds in \dev{} to yield not only higher target responses, but also higher non-target responses. To resolve this apparent paradox, we will turn to an analysis of the spatial arrangement of activity and adaptation.

@@

Finally, we examined the cause of the reduced response in non-target trials. Going from the \msc{} sequence to the \dev{} sequence, two things change: Firstly, stimulations with non-target C, D and E are replaced with B, which would cause different responses even in the absence of adaptation. Secondly, the resulting  frequent presentation of B changes TA values, likely causing adaptation to B and thereby decreasing the average response. We hypothesized that this adaptation is mediated by a small subset of neurons that (1) responds very quickly and rather strongly to stimulation in B, and therefore more to non-target stimuli in \dev{} than \msc{}, and (2) has increased TA levels as a result, reducing the response of the remaining downstream neurons in the same way as demonstrated in reverse in Figure \ref{fig4}.

@@

A measure of STD is necessarily activity-dependent -- a postsynaptic neuron's membrane potential is driven only by presynaptic partners that fire -- so it must be calculated based on observed activity. In order to respect this notion while at the same time minimizing the influence of activity differences between sequences, we used the average response $\Rhat{S}{j} = \frac{1}{3} (\R[dev]{S}{j} + \R[msc]{S}{j} + \R[std]{S}{j})$ of a neuron $j$ to a given stimulus $S$ as a measure of the presynaptic drive. Before depression, a single excitatory presynaptic spike evokes a postsynaptic potential (EPSP) peaking at approximately $k = 1.4$ mV at $V_m = V_{rest}$ (recall that $w = 1$ for all synapses), and STD linearly scales this EPSP by presynaptic $x_j$. Thus, we estimated the average peak EPSP evoked by a spike in presynaptic neuron $j$ as $\Rhat{S}{j} x_j^S k$, where we ignored added depression due to repeated spikes within a trial and instead defined $x_j^S$ as the mean depression at the start of trials. Finally, summing over presynaptic neurons, we estimated the peak EPSP lost to depression in each neuron $i$ as
\begin{equation}
    \D{S}{i} = \sum_{j \in pre} \Rhat{S}{j} (1-x_j^S) k \label{eqn-D}
\end{equation}
Notice that $\D{}{}$ is measured in volts due to scaling with $k$, and is therefore roughly comparable to $\T{}{}$, albeit more approximate.

>>>>

Having established that the increased response to A in \dev{} trials was driven by lower thresholds before stimulation, we next turned to the cause of this reduction. A priori, since threshold adaptation is a direct reflection of activity, and since the response to target trials was larger in \dev{} than in \msc{}, the lower thresholds in \dev{} must have been caused by lower activity in non-target trials, i.e., by lower responses to B in \dev{} than to B through E in \msc{}. Yet, since activity was modulated by thresholds as shown above, and since the dev sequence was random, implying that $\T[dev]{A}{} \approx \T[dev]{B}{}$, we might expect the lowered thresholds in \dev{} to yield not only higher target responses, but also higher non-target responses. To resolve this apparent paradox, we will turn to an analysis of the spatial arrangement of activity and adaptation.

@@

Finally, we examined the cause of the reduced response in non-target trials. Going from the \msc{} sequence to the \dev{} sequence, two things change: Firstly, stimulations with non-target C, D and E are replaced with B, which would cause different responses even in the absence of adaptation. Secondly, the resulting  frequent presentation of B changes TA values, likely causing adaptation to B and thereby decreasing the average response. We hypothesized that this adaptation is mediated by a small subset of neurons that (1) responds very quickly and rather strongly to stimulation in B (and therefore more to non-target stimuli in \dev{} than \msc{}), and (2) has increased TA levels as a result, reducing the response of the remaining downstream neurons in the same way as demonstrated in reverse in Figure \ref{fig4}.

@@

A measure of STD is necessarily activity-dependent -- a postsynaptic neuron's membrane potential is driven only by presynaptic partners that fire -- so it must be calculated based on observed activity. In order to respect this notion while at the same time minimizing the influence of activity differences between sequences, we used the average response
\begin{equation}
    \Rhat{S}{j} = \frac{1}{3} (\R[dev]{S}{j} + \R[msc]{S}{j} + \R[std]{S}{j}) \label{eqn-Rhat}
\end{equation}
of a neuron $j$ to a given stimulus $S$ as a measure of the presynaptic drive. Before depression, a single excitatory presynaptic spike evokes a postsynaptic potential (EPSP) peaking at approximately $k = 1.4$ mV at $V_m = V_{rest}$ (recall that $w = 1$ for all synapses), and STD linearly scales this EPSP by presynaptic $x_j$. Thus, we estimated the average peak EPSP evoked by a spike in presynaptic neuron $j$ as $\Rhat{S}{j} x_j^S k$, where we ignored added depression due to repeated spikes within a trial and instead defined $x_j^S$ as the mean depression at the start of trials. Finally, summing over presynaptic neurons, we estimated the peak EPSP lost to depression in each neuron $i$ as
\begin{equation}
    \D{S}{i} = \sum_{j \in pre} \Rhat{S}{j} (1-x_j^S) k \label{eqn-D}
\end{equation}
Notice that $\D{}{}$ is measured in volts due to scaling with $k$, and is therefore roughly comparable to $\T{}{}$, albeit more approximate.
